\huge UNet-HCRF Integration for EEG Pattern Recognition

### 2024-12-12 manuscript/main.tex
edit abs, res

%In this project, we will integrate powerful deep neural networks and
%statistical methods with a probabilistic graphical model (PGM) to effectively
%tackle sequence labeling tasks on a heterogeneous medical care dataset.

@@

In recent years, deep learning has emerged as a powerful tool to automate and
improve EEG pattern recognition~\citep{craik2019deep}. Neural network
architectures like convolutional neural network (CNN), long short-term memory
(LSTM) and Transformer models have revolutionized EEG data processing.
Additionally, hidden conditional random fields (HCRFs) have been employed for
modeling label dependencies, demonstrating promising results.
While these methods have shown individual strengths, a comprehensive
approach that leverages their complementary capabilities is essential for
tackling complex, multi-modal datasets.

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{workframe}
\caption{The proposed feature-learning-based HCRF using UNet feature maps.}
\label{fig:workframe}
\end{figure*}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=.48\textwidth]{example}
\caption{Examples of EEG patterns with different levels of expert agreement,
sourced from the HMS dataset.}
\label{fig:ex}
\end{figure}

@@

\begin{figure}[bp]
\centering
\includegraphics[width=.45\textwidth]{inion}
\caption{Chains employed to determine representative montage features via
a bipolar montage, sourced from Kaggle.}
\label{fig:inion}
\end{figure}

@@

\section{Results}

\section{Conclusion}

>>>>

Electroencephalography (EEG) pattern recognition presents critical challenges
in neurological diagnostics, where precise classification of brain activity can
significantly impact patient outcomes. The complexity of EEG signals pose
substantial challenges for traditional modeling techniques. This research
introduces a novel framework integrating deep neural networks and probabilistic
graphical models to enhance EEG pattern recognition performance. The proposed
approach combines UNet-based feature extraction with hidden conditional random
fields (HCRFs) to improve brain activity classification accuracy.
The methodology employs advanced preprocessing techniques, including wavelet
denoising and spectrogram transformation of EEG signals from the Harmful Brain
Activity Classification dataset, which contains 50-second brain activity
samples labeled into six distinct categories. The research addresses challenges
in neurological signal processing by developing a sophisticated model capable
of capturing intricate temporal dependencies and handling varying levels of
expert annotation agreement. Experimental results demonstrated significant
improvements in classifying complex EEG patterns across different levels of
expert consensus. The complete implementation code is available on GitHub at
\url{https://github.com/Carol-seven/UNet-HCRF_EEG}.

@@

In recent years, deep learning has emerged as a powerful tool to automate and
improve EEG pattern recognition~\citep{craik2019deep}. Neural network
architectures like convolutional neural network (CNN), long short-term memory
(LSTM) and Transformer models have revolutionized EEG data processing.
Additionally, conditional random fields (CRFs) have been employed for
modeling label dependencies, demonstrating promising results.
While these methods have shown individual strengths, a comprehensive
approach that leverages their complementary capabilities is essential for
tackling complex, multi-modal datasets.

@@

\begin{figure*}[tp]
\centering
\includegraphics[width=\textwidth]{workframe}
\caption{The proposed feature-learning-based HCRF using UNet feature maps.}
\label{fig:workframe}
\end{figure*}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=.48\textwidth]{example}
\caption{Examples of EEG patterns with different levels of expert agreement,
sourced from the HMS dataset.}
\label{fig:ex}
\end{figure}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=.45\textwidth]{inion}
\caption{Chains employed to determine representative montage features via
a bipolar montage, sourced from Kaggle.}
\label{fig:inion}
\end{figure}

@@

\section{Results and Discussion}

In this study, we developed a novel UNet-HCRF model for EEG pattern recognition.
Our model achieved a competitive KL divergence of 0.4992 on a five-fold
cross-validation, demonstrating competitive performance compared to the
baseline EfficientNet model from Kaggle, which had a KL divergence of 0.59.
However, our results indicated room for further improvement, as state-of-the-art
models such as Vision Transformers (ViT) have achieved significantly lower KL
divergence scores, ranging between 0.2 and 0.3.

The proposed UNet-HCRF framework integrates a latent CRF layer with a neural
network architecture to capture intricate segmentation information from EEG
dataset. By employing a variational mean-field inference scheme and an
end-to-end training strategy, we efficiently learned both neural network
features and the CRF parameters. This approach offers a flexible and powerful
framework for EEG analysis, leveraging the strengths of both deep learning and
probabilistic graphical models.

Our findings demonstrate the potential of HCRF models for EEG pattern
classification. However, several avenues for future research remain. For
instance, exploring alternative variational inference techniques or
incorporating attention mechanisms within the CRF layer could potentially
enhance model performance. Additionally, investigating the impact of different
neural network architectures and hyperparameter tuning on the overall model
performance is also a worthwhile direction for further research.
